---
title: "01 - Set Up Cannabis  Heart Rate Data For Analysis"
author: "Harry Smith"
date: "02/25/2022"
output: html_document
---

Data was received from Nichole Reisdorph on April 7, 2021 in two excel files:

"/Users/smithh/Documents/Reisdorph_Lab/Reisdorph_Cannibus_Study/Laura_dropbox/cannabis/data/July_BPM_Data.csv"

"/Users/smithh/Documents/Reisdorph_Lab/Reisdorph_Cannibus_Study/Laura_dropbox/cannabis/data"

Data needs to be reformatted to enable longitudinal analyses in R.

```{r setup, include=FALSE, echo=FALSE}
rm(list=ls())

options(dplyr.summarise.inform=F)


library(dplyr)
library(tidyr)
library(kableExtra)

wd <- "/Users/smithh/Documents/Reisdorph_Lab/Reisdorph_Cannibus_Study/Laura_dropbox/cannabis/data/"
```

## bpm data from July

```{r, echo=FALSE}
july = read.table(file=paste0(wd,"July_BPM_Data.csv"), sep=",", header=TRUE, fill=TRUE)
july_t = july %>% 
  pivot_longer(c(5:16), names_to = "sample_id", values_to = "bpm") %>%
  mutate(mouse = gsub("X","",unlist(lapply(strsplit(sample_id, split=".", fixed=TRUE), function(a) a[1]))),
         admin = unlist(lapply(strsplit(sample_id, split=".", fixed=TRUE), function(a) a[2])),
         dose = unlist(lapply(strsplit(sample_id, split=".", fixed=TRUE), function(a) a[length(a)])))

july_t$dose[july_t$dose=="high"] = "High"
july_t$dose[july_t$dose=="low"] = "Low"

july_t$day_type = c("dose", "no dose")[as.numeric(grepl("No",july_t$time_period)) + 1]
  
  
mice = july_t %>%
  group_by(mouse, admin, dose) %>%
  summarize(n_obs = sum(!is.na(bpm)))

july_timeline = july_t %>%
  group_by(date,study_day,time_period) %>%
  summarize(n_obs = sum(!is.na(bpm)))

july_timeline = july_timeline[order(july_timeline$study_day),]
colnames(july_timeline) = c("Date", "Study Day", "Type of Day", "Number of Measured bpm")
```

Number of mice: `r nrow(mice)`

Total number of bpmerature measurements: `r prettyNum(sum(!is.na(july_t$bpm)), big.mark=",")`

Range of number of bpmerature measurements per mouse: `r min(mice$n_obs)` to `r max(mice$n_obs)`

```{r, echo=FALSE}
hist(july_t$bpm, main="", xlab="Heart Rate - BPM")
```

```{r, echo=FALSE}
july_table = mice %>%
  group_by(admin, dose) %>%
  summarize(num_mice = length(mouse))
colnames(july_table) = c("Administration Type","Dose Level","Number of Mice")
```

**Treatment Group Assignment**

```{r, results='asis',echo=FALSE, eval=TRUE}
kable(july_table, "html",align=rep("c",ncol(july_table)), row.names = FALSE) %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position="left")
```

**TimeLine**

```{r, results='asis',echo=FALSE, eval=TRUE}
kable(july_timeline, "html",align=rep("c",ncol(july_timeline)), row.names = FALSE) %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position="left")
```


## bpm data from January

```{r, echo=FALSE}
jan = read.table(file=paste0(wd,"January_BPM_Data.csv"), sep=",", header=TRUE, fill=TRUE)
jan_t = jan %>% 
  pivot_longer(c(5:15), names_to = "sample_id", values_to = "bpm") %>%
  mutate(mouse = gsub("X","",unlist(lapply(strsplit(sample_id, split=".", fixed=TRUE), function(a) a[1]))),
         admin = unlist(lapply(strsplit(sample_id, split=".", fixed=TRUE), function(a) a[2])),
         dose = unlist(lapply(strsplit(sample_id, split=".", fixed=TRUE), function(a) a[length(a)])))

jan_t$admin[jan_t$admin=="Ip"] = "IP"
jan_t = jan_t[jan_t$time_period!="Sac Day",]  ##need to double check with Nichole about this

jan_t$day_type = c("dose", "no dose")[as.numeric(grepl("No",jan_t$time_period)) + 1]

jan_mice = jan_t %>%
  group_by(mouse, admin, dose) %>%
  summarize(n_obs = sum(!is.na(bpm)))

jan_timeline = jan_t %>%
  group_by(date,study_day,time_period) %>%
  summarize(n_obs = sum(!is.na(bpm)))

jan_timeline = jan_timeline[order(jan_timeline$study_day),]
colnames(jan_timeline) = c("Date", "Study Day", "Type of Day", "Number of Measured bpm")

```

Number of mice: `r nrow(jan_mice)`

Total number of bpmerature measurements: `r prettyNum(sum(!is.na(jan_t$bpm)), big.mark=",")`

Range of number of bpmerature measurements per mouse: `r min(jan_mice$n_obs)` to `r max(mice$n_obs)`

```{r, echo=FALSE}
hist(jan_t$bpm, main="", xlab="Heart Rate - BPM")
```

```{r, echo=FALSE}
jan_table = jan_mice %>%
  group_by(admin, dose) %>%
  summarize(num_mice = length(mouse))
colnames(jan_table) = c("Administration Type","Dose Level","Number of Mice")
```

**Treatment Group Assignment**

```{r, results='asis',echo=FALSE, eval=TRUE}
kable(jan_timeline, "html",align=rep("c",ncol(jan_timeline)), row.names = FALSE) %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position="left")
```

**Timeline**

```{r, results='asis',echo=FALSE, eval=TRUE}
kable(jan_table, "html",align=rep("c",ncol(jan_table)), row.names = FALSE) %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position="left")
```

```{r, echo=FALSE, eval=FALSE}
Rdata_dir <- "/Users/smithh/Documents/Reisdorph_Lab/Reisdorph_Cannibus_Study/Laura_dropbox/cannabis/Rdata/"
save(jan_t, july_t, file=paste0(Rdata_dir,"bpm_data.Rdata"))
```