---
title: "02a - Longitudinal Analysis of Cannabis-related Heart Rate (Free Feed Administration)"
author: "Harry Smith"
date: "`r Sys.Date()`"
output: html_document
---



```{r setup, include=FALSE, echo=FALSE}
rm(list=ls())

options(dplyr.summarise.inform=F)

library(dplyr)
library(tidyr)
library(kableExtra)
library(lmerTest)
library(emmeans)
library(pbkrtest)
library(ggplot2)
library(longpower)
library(gee)


Rdata_dir <- "/Users/smithh/Documents/Reisdorph_Lab/Reisdorph_Cannibus_Study/Laura_dropbox/cannabis/Rdata/"
```

## Free Feed Administration Model - July

```{r echo=FALSE}
load(paste0(Rdata_dir,"bpm_data.Rdata"))

free_feed <- july_t %>% filter(admin=="Free")
free_feed$round <- ceiling(free_feed$study_day/2)
free_feed$day_type = factor(free_feed$day_type)
free_feed$time_window = factor(free_feed$time_window,
                               levels=c( "PreDose","Dose","Post Dose 1","Post Dose 2","Post Dose 3","Post Dose 4","Post Dose 5","Post Dose 6"))
free_feed$dose = factor(free_feed$dose, levels=c("Control","Low","Med","High"))
free_feed$round = factor(free_feed$round)

ff_model = lmer(bpm ~ time_window*dose*round+ (1|mouse) , data=free_feed[free_feed$day_type=="dose",])

# ## Power analysis for Nicholes new grant
# 
# # first run gee test
# tempstudy <- free_feed[free_feed$day_type=="dose",]
# fm3 <- lme(temp ~ time_window*dose*round, random=~1|mouse, tempstudy)
# 
# # power calc
# pwr_res <- lmmpower(x, 
#          sig.level = 0.05, 
#          power = .80, 
#          pct.change = .25, 
#          parameter = 2, 
#          t = c(0,1,2,3,4,5,6))
# 
# powerSim(fm3)
```

## Data Dictionary

<style type="text/css">
.tg  {border-collapse:collapse;border-color:#ccc;border-spacing:0;}
.tg td{background-color:#fff;border-color:#ccc;border-style:solid;border-width:1px;color:#333;
  font-family:Arial, sans-serif;font-size:14px;overflow:hidden;padding:10px 5px;word-break:normal;}
.tg th{background-color:#f0f0f0;border-color:#ccc;border-style:solid;border-width:1px;color:#333;
  font-family:Arial, sans-serif;font-size:14px;font-weight:normal;overflow:hidden;padding:10px 5px;word-break:normal;}
.tg .tg-0pky{border-color:inherit;text-align:left;vertical-align:top}
.tg .tg-0lax{text-align:left;vertical-align:top}
</style>
<table class="tg">
<thead>
  <tr>
    <th class="tg-0pky">Variable</th>
    <th class="tg-0pky">Definition</th>
    <th class="tg-0pky">Factor Levels</th>
  </tr>
</thead>
<tbody>
  <tr>
    <td class="tg-0pky">date</td>
    <td class="tg-0pky">The actual date associated with record</td>
    <td class="tg-0pky">N/A</td>
  </tr>
  <tr>
    <td class="tg-0pky">study_day</td>
    <td class="tg-0pky">An ordinal number describing the study day. For each study day, there are 72 records</td>
    <td class="tg-0pky">N/A</td>
  </tr>
  <tr>
    <td class="tg-0pky">time_period</td>
    <td class="tg-0pky">Variable describing the dose period. Each time period can be either a "dose #" or a "nonDose #". Example: Dose 1 &amp; nonDose 1</td>
    <td class="tg-0pky">N/A</td>
  </tr>
  <tr>
    <td class="tg-0lax">time_window</td>
    <td class="tg-0lax">Variable describing aggregated time windows for temp record collection. Original data had temp data recorded <br>by the minute. These data were then averaged over 6 minute windows, and then aggregated in to hour long windows.<br>There 8 windows</td>
    <td class="tg-0lax">pre dose<br>dose<br>post dose 1<br>post dose 2<br>post dose 3<br>post dose 4<br>post dose 5<br>post dose 6</td>
  </tr>
  <tr>
    <td class="tg-0lax">sample_id</td>
    <td class="tg-0lax">Unique identifier for each mouse</td>
    <td class="tg-0lax">NA</td>
  </tr>
  <tr>
    <td class="tg-0lax">temp</td>
    <td class="tg-0lax">Aggregated heart rate measurement for a given time_window</td>
    <td class="tg-0lax">N/A</td>
  </tr>
  <tr>
    <td class="tg-0lax">mouse</td>
    <td class="tg-0lax">Number assigned to a given mouse</td>
    <td class="tg-0lax">N/A</td>
  </tr>
  <tr>
    <td class="tg-0lax">admin</td>
    <td class="tg-0lax">Method used to administer cannabis to mouse</td>
    <td class="tg-0lax">Free<br>IP<br>Gavage</td>
  </tr>
  <tr>
    <td class="tg-0lax">dose</td>
    <td class="tg-0lax">Variable describing the dose level for a given mouse</td>
    <td class="tg-0lax">Control<br>Low<br>Medium<br>High</td>
  </tr>
  <tr>
    <td class="tg-0lax">day_type</td>
    <td class="tg-0lax">Variable describing whether or not a dose of cannabis was administered</td>
    <td class="tg-0lax">dose<br>no dose</td>
  </tr>
  <tr>
    <td class="tg-0lax">round</td>
    <td class="tg-0lax">Variable that serves as a wrapper for an administration cycle (i.e. encompasses a dose day and a no dose day)</td>
    <td class="tg-0lax">N/A</td>
  </tr>
</tbody>
</table>

### Chronic effects of cannabis

For this analysis, I compared the heart rate at during the 'predose' window across dosage groups using time in days as a continuous covariate. Data from dosing days and non-dosing days were included since we are only looking at the predose window.


```{r, echo=FALSE}
## chronic effects of cannabis

# difference in predose temperature across time and dose

x <- lmer(bpm ~ dose*study_day + (1|mouse),
          data=free_feed[free_feed$time_window=="PreDose",])

# # power calc
# pwr_res <- lmmpower(x, 
#          sig.level = 0.05, 
#          power = .80, 
#          pct.change = .25, 
#          parameter = 2, 
#          t = c(0,1,2,3,4,5,6))

x_trends <- emtrends(x, pairwise ~ dose, var = "study_day")

x_rg = ref_grid(x, at = list(study_day = c(1,8,14)))
x_means <- emmeans(x_rg, ~ dose, by="study_day")
x_contrasts <- contrast(x_means,method="trt.vs.ctrl", ref="Control", adjust="none")
```

**Two-Way ANOVA Results**
```{r, results='asis',echo=FALSE, eval=TRUE}
kable(anova(x), "html",align=rep("c",ncol(anova(x))), row.names = TRUE) %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position="left")
```

**Comparing Slopes, i.e., study day trends, across dosage groups**
```{r, results='asis',echo=FALSE, eval=TRUE}
kable(x_trends$emtrends, "html",align=rep("c",ncol(data.frame(x_trends$emtrends))), row.names = FALSE) %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position="left")
```


```{r, results='asis',echo=FALSE, eval=TRUE}
kable(data.frame(x_trends$contrasts), "html",align=rep("c",ncol(data.frame(x_trends$contrasts))), row.names = FALSE) %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position="left")
```

```{r, echo=FALSE}
emmip(x, dose ~ study_day, cov.reduce = range,
      xlab="Day of Study", ylab="Heart Rate During Predose Window")
```

Based on my understanding of the study design, the 4 dosage groups should not differ in heart rate at day 1 during the predose window because none of the animals would have received any cannabis prior to that time point. 

Since this graphic made it look like the dosage groups differed at day 1, I tested that directly.

```{r, echo=FALSE}
x <- lm(bpm ~ dose, data=free_feed[free_feed$study_day==1 & free_feed$time_window=="PreDose",])
x_means <- emmeans(x, ~ dose)
x_contrasts <- contrast(x_means,method="pairwise")
```

Omnibus p-value for dosage effect: `r sprintf("%.3f", round(anova(x)[[5]][1],3))`

```{r, echo=FALSE}
plot(as.numeric(free_feed$dose[free_feed$study_day==1 & free_feed$time_window=="PreDose"]), free_feed$temp[free_feed$study_day==1 & free_feed$time_window=="PreDose"], xaxt="n", pch=19,
     xlab = "Dosage Group", ylab="Heart Rate")
axis(1, at=c(1,3,5,7), labels = levels(free_feed$dose), tick=FALSE)
```    

Although there isn't a significant difference in heart rate on study day 1 at the predose time window. There is an obvious trend.

## Acute Effects of Cannabis on Temperature

To examine the acute effects (i.e., within a treatment day) of cannabis, data points were limited to those collected during a 'dose' day.

### Initial Acute Effect of Cannabis on Temperature

First, the initial acute effects of cannabis were assessed by only focusing on changes in heart rate across different windows of time on the first study day that mice were exposed to cannabis

```{r, echo=FALSE, eval=TRUE}
## initial acute effects of cannabis
x <- lmer(bpm ~ dose*time_window + (1|mouse),
          data=free_feed[free_feed$round==1,])
```

**Two-Way ANOVA Results**
```{r, results='asis',echo=FALSE, eval=TRUE}
kable(anova(x), "html",align=rep("c",ncol(anova(x))), row.names = TRUE) %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position="left")
```

There were no differences in heart rate across the different time windows within the first day of treatment.


### Acute Effect of Cannabis on Temperature After Chronic Exposure to Cannabis


```{r echo=FALSE, eval=TRUE, include=FALSE}
## acute effects of cannabis after chronic exposure
x <- lmer(bpm ~ dose*time_window + (1|mouse),
          data=free_feed[free_feed$round==7,])
x_means <- emmeans(x, ~ time_window, by="dose")
x_contrasts <- contrast(x_means,method="trt.vs.ctrl", ref="PreDose", adjust="none")

x1_means <- emmeans(x, ~ time_window)
x1_contrasts <- contrast(x1_means,method="trt.vs.ctrl", ref="PreDose", adjust="none")
```


**Two-Way ANOVA Results**
```{r, results='asis',echo=FALSE, eval=TRUE}
kable(anova(x), "html",align=rep("c",ncol(anova(x))), row.names = TRUE) %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position="left")
```

The two-way ANOVA results suggest that there is a nominal time window effect but that effect does not differ between dosage groups (i.e., the interaction effect is not significant). Therefore the time window estimates were calculated both within each dosage group and between dosage groups.

**Time Window Effect in Last Round of Dose Stratified by Dose**

```{r, echo=FALSE}
p <- emmip(x, dose ~ time_window, ylab = "heart rate in final dose round",
      xlab = "")
p + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

The only time window contrast between PreDose and each of the other time windows that was significant (unadjusted p-value < 0.05) was Post dose 4.

**Marginal Means Estimates for Time Window Effect in Last Round of Dose**

```{r, echo=FALSE, include=FALSE}
p <- emmip(x1_means,  ~ time_window, ylab = "heart rate in final dose round",
      xlab = "")
p + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

