---
title: "02a - Longitudinal Analysis of Cannabis-related Temperatures (Free Feed Administration)"
author: "Laura Saba"
date: "`r Sys.Date()`"
output: html_document
---



```{r setup, include=FALSE, echo=FALSE}
rm(list=ls())

options(dplyr.summarise.inform=F)

library(dplyr)
library(tidyr)
library(kableExtra)
library(lmerTest)
library(emmeans)
library(ggplot2)

wd <- "/Users/sabal/Documents/Collaborators/Reisdorph_Nichole/cannabis/"
```

## Free Feed Administration Model - July

```{r echo=FALSE}
load(paste0(wd,"Rdata/temp_data.Rdata"))

free_feed <- july_t %>% filter(admin=="Free")
free_feed$round <- ceiling(free_feed$study_day/2)
free_feed$day_type = factor(free_feed$day_type)
free_feed$time_window = factor(free_feed$time_window,
                               levels=c( "PreDose","Dose","Post Dose 1","Post Dose 2","Post Dose 3","Post Dose 4","Post Dose 5","Post Dose 6"))
free_feed$dose = factor(free_feed$dose, levels=c("Control","Low","Med","High"))
free_feed$round = factor(free_feed$round)

ff_model = lmer(temp ~ time_window*dose*round + (1|mouse) , data=free_feed[free_feed$day_type=="dose",])
```

### Chronic effects of cannabis

For this analysis, I compared the temperature at during the 'predose' window across dosage groups using time in days as a continuous covariate. Data from dosing days and non-dosing days were included since we are only looking at the predose window.


```{r, echo=FALSE}
## chronic effects of cannabis

# difference in predose temperature across time and dose

x <- lmer(temp ~ dose*study_day + (1|mouse),
          data=free_feed[free_feed$time_window=="PreDose",])

x_trends <- emtrends(x, pairwise ~ dose, var = "study_day")

x_rg = ref_grid(x, at = list(study_day = c(1,8,14)))
x_means <- emmeans(x_rg, ~ dose, by="study_day")
x_contrasts <- contrast(x_means,method="trt.vs.ctrl", ref="Control", adjust="none")
```

**Two-Way ANOVA Results**
```{r, results='asis',echo=FALSE, eval=TRUE}
kable(anova(x), "html",align=rep("c",ncol(anova(x))), row.names = TRUE) %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position="left")
```

**Comparing Slopes, i.e., study day trends, across dosage groups**
```{r, results='asis',echo=FALSE, eval=TRUE}
kable(x_trends$emtrends, "html",align=rep("c",ncol(data.frame(x_trends$emtrends))), row.names = FALSE) %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position="left")
```


```{r, results='asis',echo=FALSE, eval=TRUE}
kable(data.frame(x_trends$contrasts), "html",align=rep("c",ncol(data.frame(x_trends$contrasts))), row.names = FALSE) %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position="left")
```

```{r, echo=FALSE}
emmip(x, dose ~ study_day, cov.reduce = range,
      xlab="Day of Study", ylab="Temperature During Predose Window")
```

Based on my understanding of the study design, the 4 dosage groups should not differ in temperature at day 1 during the predose window because none of the animals would have received any cannabis prior to that time point. 

Since this graphic made it look like the dosage groups differed at day 1, I tested that directly.

```{r, echo=FALSE}
x <- lm(temp ~ dose, data=free_feed[free_feed$study_day==1 & free_feed$time_window=="PreDose",])
x_means <- emmeans(x, ~ dose)
x_contrasts <- contrast(x_means,method="pairwise")
```

Omnibus p-value for dosage effect: `r sprintf("%.3f", round(anova(x)[[5]][1],3))`

```{r, echo=FALSE}
plot(as.numeric(free_feed$dose[free_feed$study_day==1 & free_feed$time_window=="PreDose"]), free_feed$temp[free_feed$study_day==1 & free_feed$time_window=="PreDose"], xaxt="n", pch=19,
     xlab = "Dosage Group", ylab="Temperature")
axis(1, at=c(1:4), labels = levels(free_feed$dose), tick=FALSE)
```    

Although there isn't a significant difference in temperature on study day 1 at the predose time window. There is an obvious trend.

## Acute Effects of Cannabis on Temperature

To examine the acute effects (i.e., within a treatment day) of cannabis, data points were limited to those collected during a 'dose' day.

### Initial Acute Effect of Cannabis on Temperature

First, the initial acute effects of cannabis were assessed by only focusing on changes in temperature across different windows of time on the first study day that mice were exposed to cannabis

```{r, echo=FALSE, eval=TRUE}
## initial acute effects of cannabis
x <- lmer(temp ~ dose*time_window + (1|mouse),
          data=free_feed[free_feed$round==1,])
```

**Two-Way ANOVA Results**
```{r, results='asis',echo=FALSE, eval=TRUE}
kable(anova(x), "html",align=rep("c",ncol(anova(x))), row.names = TRUE) %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position="left")
```

There were no differences in temperature across the different time windows within the first day of treatment.


### Acute Effect of Cannabis on Temperature After Chronic Exposure to Cannabis


```{r echo=FALSE, eval=TRUE, include=FALSE}
## acute effects of cannabis after chronic exposure
x <- lmer(temp ~ dose*time_window + (1|mouse),
          data=free_feed[free_feed$round==7,])
x_means <- emmeans(x, ~ time_window, by="dose")
x_contrasts <- contrast(x_means,method="trt.vs.ctrl", ref="PreDose", adjust="none")

x1_means <- emmeans(x, ~ time_window)
x1_contrasts <- contrast(x1_means,method="trt.vs.ctrl", ref="PreDose", adjust="none")
```


**Two-Way ANOVA Results**
```{r, results='asis',echo=FALSE, eval=TRUE}
kable(anova(x), "html",align=rep("c",ncol(anova(x))), row.names = TRUE) %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position="left")
```

The two-way ANOVA results suggest that there is a time window effect but that effect does not differ between dosage groups (i.e., the interaction effect is not significant). Therefore the time window estimates were calculated both within each dosage group and between dosage groups.

**Time Window Effect in Last Round of Dose Stratified by Dose**

```{r, echo=FALSE}
p <- emmip(x, dose ~ time_window, ylab = "temperature in final dose round",
      xlab = "")
p + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

None of the pairwise contrasts between PreDose and each of the other time windows were significant (unadjusted p-value < 0.05) for any of the different dosage groups.

**Marginal Means Estimates for Time Window Effect in Last Round of Dose**

```{r, echo=FALSE, include=FALSE}
p <- emmip(x1_means,  ~ time_window, ylab = "temperature in final dose round",
      xlab = "")
p + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

None of the pairwise contrasts between PreDose and each of the other time windows were significant (unadjusted p-value < 0.05).
